---
title: "gapminder_exploration"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Bring the rectangular data in
```{r}
library(gapminder)
library(tidyverse)
```


##Small test the data
1. Is it a data.frame, a matrix,a vector, a list?
Answer: As we called the str() function, we can clearly see that our data is a data.frame. 
```{r}
str(gapminder)
typeof(gapminder)
```

2. What is its class?
As shown below, the class() function returned the class of the data, which are table, tibble and data frame. 

```{r}
class(gapminder) ##class
```

3. How many variables/column?
There are 6 columns when each column corresponding to one variable.
```{r}
ncol(gapminder) ## number of column
```


4. How mant rows/observation?
There are total 1704 rows which stands for 1704 observations. 
```{r}
nrow(gapminder)
```

5. Get these facts about extent or size in more than one way?
```{r}
length(gapminder) ## return the length of the vectors therefore the number of column
dim(gapminder)  ## return the dimension of the vector in the format of row by column, therefore, 1704 rows and 6 column for this database

nrow <- dim(gapminder)[1] ## return number of rows
ncol <- dim(gapminder)[2] ## return number of columns
```

6.What data type is each variable?

|variable| data type|
-------------------
|--------|----------|
|  country   |  integer  |
|   gdpPercap  | double  |
| continent  |  integer  |
|  year    | integr  |
|  pop   | integer  |
|  lifeExp  | double  |



```{r}
typeof(gapminder$country)
typeof(gapminder$gdpPercap)
typeof(gapminder$continent)
typeof(gapminder$year)
typeof(gapminder$pop)
typeof(gapminder$lifeExp)
```


## Exploring individual varaibles
Categorical variable:continent
Quantitative variable:year
```{r}
summary(gapminder)
summary(gapminder$continent) ##sorting the possible continent and also the number of country each continent contains

max(gapminder$year)
min(gapminder$year)
var(gapminder$year)
```
As we can explore from the summary, for variable Continent, the possible values are Africa, Americas, Asia, Europe and Oceania. As for the quantitative variable year, the range of variable is from 1952 to 2007.

```{r}
summary(gapminder$year)
```
As shown from the above data, although year is ranging from 1952-2007, the most typical values are ranging from 1966 to 1993 with a median and mean of 1980.t The variance of this variable is 298.0916.

For the variable continent, the most typical value is Africa and the continent Oceania has less entries. 


## Explore various plot types
This is the scatterplot of pop vs gdpPercap 
```{r}
ggplot(gapminder,aes(pop,gdpPercap))+
  geom_point(alpha=0.25)
```

This is a histogram of pop and the bin is set to be 50.
```{r}
ggplot(gapminder,aes(pop))+
  geom_histogram(bins=50)
```
This is a combination of kernel density plot and histogram of pop and the bin is set to be 50.

```{r}
ggplot(gapminder,aes(pop))+
  geom_histogram(aes(y=..density..))+
  geom_density()
```
This is a boxplot of lifeExp for each continent.
```{r}
ggplot(gapminder,aes(continent,lifeExp))+
  geom_boxplot()
```

This is the violin plot of lifeExp for each continent
```{r}
ggplot(gapminder,aes(continent,lifeExp))+
  geom_violin()
```

##Using filter(), select() and %>% 

filter the data to continent Oceania only and the various plots are shown below:

```{r}
gapminder %>% 
  filter(continent=="Oceania") %>% 
  ggplot(aes(year, pop))+
 geom_point()
```

select only variables lifeExp and continent and plot the boxplot and jitter plot of lifeExp for each continent.
```{r}
gapminder %>% 
 select(lifeExp,continent) %>% 
  ggplot(aes(continent,lifeExp))+
  geom_boxplot(alpha=0.25)+
  geom_jitter()

```

## Extra Work
```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```
This is the correct way to extract the entries for country Afghanistan and Rwanda only.
This only filter out the entry that country=="Rwanda" and country=="Afghanistan"



