---
title: "gapminder_exploration"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Bring the rectangular data in
```{r}
library(gapminder)
library(tidyverse)
```


##Small test the data
1. Is it a data.frame, a matrix,a vector, a list?
Answer: As we called the str() function, we can clearly see that our data is a data.frame. 
```{r}
str(gapminder)
typeof(gapminder)
```

2. What is its class?
As shown below, the class() function returned the class of the data, which are table, tibble and data frame. 

```{r}
class(gapminder) ##class
```

3. How many variables/column?
There are 6 columns when each column corresponding to one variable.
```{r}
ncol(gapminder) ## number of column
```


4. How mant rows/observation?
There are total 1704 rows which stands for 1704 observations. 
```{r}
nrow(gapminder)
```

5. Get these facts about extent or size in more than one way?
```{r}
length(gapminder) ## return the length of the vectors therefore the number of column
dim(gapminder)  ## return the dimension of the vector in the format of row by column, therefore, 1704 rows and 6 column for this database

nrow <- dim(gapminder)[1] ## return number of rows
ncol <- dim(gapminder)[2] ## return number of columns
```

6.What data type is each variable?
Here are the data type for each variable:

```{r}
typeof(gapminder$country)
typeof(gapminder$gdpPercap)
typeof(gapminder$continent)
typeof(gapminder$year)
typeof(gapminder$pop)
typeof(gapminder$lifeExp)
```


## Exploring individual varaibles
Categorical variable:continent
Quantitative variable:year
```{r}
summary(gapminder)  ## this has summarized the data of gapminder and give us a general view of the data.
summary(gapminder$continent) ##sorting the possible continent and also the number of country each continent contains

max(gapminder$year)
min(gapminder$year)
var(gapminder$year)
```
As we can explore from the summary, for variable Continent, the possible values are Africa, Americas, Asia, Europe and Oceania. As for the quantitative variable year, the range of variable is from 1952 to 2007.

```{r}
summary(gapminder$year)
```
As shown from the above data, although year is ranging from 1952-2007, the most typical values are ranging from 1966 to 1993 with a median and mean of 1980.t The variance of this variable is 298.0916.

For the variable continent, the most typical value is Africa and the continent Oceania has less entries. 


## Explore various plot types

#This is the scatterplot of pop vs gdpPercap, transparentcy of 0.25 is applied to give a better view of the data. 
```{r}
ggplot(gapminder,aes(pop,gdpPercap))+
  geom_point(alpha=0.25)
```

#This is a histogram of pop and the bin is set to be 50.
```{r}
ggplot(gapminder,aes(pop))+
  geom_histogram(bins=50)
```
#This is a combination of kernel density plot and histogram of pop and the bin is set to be 50.

```{r}
ggplot(gapminder,aes(pop))+
  geom_histogram(aes(y=..density..))+
  geom_density()
```
#This is a boxplot of lifeExp for each continent.
```{r}
## fill is used to give different continent a different color filling 
ggplot(gapminder,aes(continent,lifeExp))+
  geom_boxplot(aes(fill=continent))
```

#This is the violin plot of lifeExp for each continent
```{r}
ggplot(gapminder,aes(continent,lifeExp))+
  geom_violin(aes(fill=continent))
```

#separate plotting of pop for each continent
Since different continent has various population, the  scales free_x is used to better present the data in the graph. 
```{r}
ggplot(gapminder,aes(pop))+
  facet_wrap(~ continent, scales="free_x")+
  geom_histogram(aes(fill=continent))
```



##Using filter(), select() and %>% 

filter the data to continent Americas only and the various plots are shown below:
the color function is applied to the plot to indicate the different trends for various separate countries in Americas.

```{r}
gapminder %>% 
  filter(continent=="Americas") %>% 
  ggplot(aes(year, pop))+
 geom_point(aes(color=country))
  
```

select only variables lifeExp and continent and plot the boxplot and jitter plot of lifeExp for each continent.
```{r}
gapminder %>% 
 select(lifeExp,continent) %>% 
  ggplot(aes(continent,lifeExp))+
  geom_boxplot(alpha=0.25, aes(fill=continent))+
  geom_jitter()

```

## Extra Work
```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```
This is the correct way to extract the entries for country Afghanistan and Rwanda only.
This only filter out the entry that country=="Rwanda" and country=="Afghanistan"



